---
title: "Time Series"
author: "Gaurika_Tyagi"
date: "June 4, 2016"
output: html_document
---

The scan() function assumes that the data for successive time
points is in a simple text file with one column. The kings dataset contains data on the age of death of successive kings of England. It ha 42 observations.

```{r}
kings_death <- scan("http://robjhyndman.com/tsdldata/misc/kings.dat",skip=3)
head(kings_death)

births <- scan("http://robjhyndman.com/tsdldata/data/nybirths.dat")

souvenir <- scan("http://robjhyndman.com/tsdldata/data/fancy.dat")
```

Time series analysis requires that the dataset is convrted to a time-series object before the analysis. By default ts() assumes that the data has been collected on an yearly basis. If the data has been collected on a monthly basis, we can set the frequency=12, for quarterly we set the frequency=4.

```{r}
death_timeseries <- ts(kings_death)
death_timeseries

birthstimeseries <- ts(births, frequency=12, start=c(1946,1))
birthstimeseries

#data is gathered monthly from January, 1987
souvenirtimeseries <- ts(souvenir, frequency=12, start=c(1987,1))
souvenirtimeseries
```